Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-11 manuscript/draft.Rmd
don't discuss excluded sites after counting the sites

Bird species richness was obtained from the Breeding Bird Survey (BBS)
[@pardieck2017] using the Data Retriever software [@morris2013] and
rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
all nocturnal, cepuscular, and aquatic species
[since these species are not well sampled by BBS methods; @hurlbert2005], as
well as unidentified species, and hybrids. All data from routes that did not
meet BBS acceptable criteria were excluded. Only routes within the continental
United States were used so as to match the climate dataset. The resulting
dataset included [[N]] species and [[N]] sites. Among sites richness varied from
[[N]] to [[N]] with an average richness of [[N]] species. Within-site richness
variation (among years) was small (~5 species).

Environmental data included a combination of elevation, bioclimatic variables
and a remotely sensed vegetation index (the normalized difference vegetation
index; NDVI), all of which are know to influence richness and distribution in
the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions [@harris2015]
for each year in the dataset: mean diurnal range, isothermality, max temperature
of the warmest month, mean temperature of the wettest quarter, mean temperature
of the driest quarter, precipitation seasonality, precipitation of the wettest
quarter, precipitation of the warmest quarter. Satellite derived NDVI, a primary
correlated of richness in BBS data [@hurlbert2002], with an 8 km resolution was
obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
period 1981-2013. Average summer (May, June, July) and winter (December,
January, Feburary) NDVI values were used as predictors. Elevation was from the
SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. All environmental variables were aggregated to the average
values within a 40 km radius (the length of a BBS route) of the starting point
of each BBS route.

Data from `r settings$start_yr` (the first complete year of NDVI data) to `r
settings$last_train_year` were used for training models and data from `r
settings$last_train_year + 1` to `r settings$end_yr` was used for testing. Only
BBS routes with at least `r settings$min_num_yrs` annual observations in the
years 1982-2015 were used.

>>>>

Bird species richness was obtained from the Breeding Bird Survey (BBS)
[@pardieck2017] using the Data Retriever software [@morris2013] and
rdataretriever R package [@mcglinn2017]. The BBS data was filtered to exclude
all nocturnal, cepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. Due to limitations of the environmental data (below), we did not
include any data from outside the continental United States. Observed richness
from 1982 (the first complete year of NDVI data) to 2003 were used for training
the models, and data from 2004 to 2013 was used for testing their performance.
Only BBS routes that were sampled during 70% of the years in the training period
were used (i.e., routes with at least 16 annual observations). All data from
routes that did not meet BBS acceptable criteria were excluded, as were routes
that had not been visited during at least 70% of the years in our model-training
window. The resulting dataset included [[N]] species and [[N]] sites. Among
sites,  richness varied from [[N]] to [[N]] with an average richness of [[N]]
species. Within-site richness variation (among years) was small (standard
deviation of about 5 species, after controlling for observer effects).

Environmental data included a combination of elevation, bioclimatic variables
and a remotely sensed vegetation index (the normalized difference vegetation
index; NDVI), all of which are known to influence richness and distribution in
the BBS data. We used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions [@harris2015]
for each year in the dataset: mean diurnal range, isothermality, max temperature
of the warmest month, mean temperature of the wettest quarter, mean temperature
of the driest quarter, precipitation seasonality, precipitation of the wettest
quarter, precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], with an 8 km resolution was
obtained from the NDIV3g dataset [@pinzon2014] and was available from the time
period 1981-2013. Average summer (May, June, July) and winter (December,
January, Feburary) NDVI values were used as predictors. Elevation was from the
SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. All environmental variables were aggregated to the average
values within a 40 km radius (the length of a BBS route) of the starting point
of each BBS route.
